---
title: "p23259 Gerardo figure requests"
author: "ENPRC Genomics Core"
date:  "`r Sys.Date()`"
output: 
  rmdformats::robobook:
    fig_width: 9
    css: "stylesheets/robobook.css"
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = TRUE,
                      message = TRUE, 
                      echo=FALSE, 
                      fig.width = 8, 
                      fig.height = 4)
```

# PBMC

## 12 v pre

### Reactome

```{r, fig.height=10}
gsea_plots[['PBMC_ind_pre-v-12_reactome']] 
```

### KEGG

```{r, fig.height=6}
gsea_plots[['PBMC_ind_pre-v-12_kegg']]
```

### Hallmark

```{r, fig.height=6}
gsea_plots[['PBMC_ind_pre-v-12_hallmark']]
```

## 18 v pre

### Reactome

```{r, fig.height=10}
gsea_plots[['PBMC_ind_pre-v-18_reactome']]
```

### KEGG

```{r, fig.height=6}
gsea_plots[['PBMC_ind_pre-v-18_kegg']]
```

### Hallmark

```{r, fig.height=6}
gsea_plots[['PBMC_ind_pre-v-18_hallmark']]
```

## 18 v 12

### Reactome

```{r, fig.height=10}
gsea_plots[['PBMC_ind_18-v-12_reactome']]
```

### KEGG

```{r, fig.height=6}
gsea_plots[['PBMC_ind_18-v-12_kegg']]
```

### Hallmark

```{r, fig.height=6}
gsea_plots[['PBMC_ind_18-v-12_hallmark']]
```

## Joint

### Reactome

```{r, fig.height=9}
gsea_plots[['PBMC_joint_reactome']]
```

### KEGG

```{r, fig.height=7}
gsea_plots[['PBMC_joint_kegg']]
```

### Hallmark

```{r, fig.height=7}
gsea_plots[['PBMC_joint_hallmark']]
```

## Running enrichment plots {.tabset}

### HALLMARK_INFLAMMATORY_RESPONSE

```{r}
gsea_plots$PBMC_HALLMARK_INFLAMMATORY_RESPONSE
```

### REACTOME_INTERFERON_GAMMA_SIGNALING

```{r}
gsea_plots$PBMC_REACTOME_INTERFERON_GAMMA_SIGNALING
```

### REACTOME_INTERFERON_SIGNALING

```{r}
gsea_plots$PBMC_REACTOME_INTERFERON_SIGNALING
```

### REACTOME_SIGNALING_BY_INTERLEUKINS

```{r}
gsea_plots$PBMC_REACTOME_SIGNALING_BY_INTERLEUKINS
```

### REACTOME_TOLL_LIKE_RECEPTOR_CASCADES

```{r}
gsea_plots$PBMC_REACTOME_TOLL_LIKE_RECEPTOR_CASCADES
```

### KEGG_B_CELL_RECEPTOR_SIGNALING_PATHWAY

```{r}
gsea_plots$PBMC_KEGG_B_CELL_RECEPTOR_SIGNALING_PATHWAY
```

# RB

## 12 v pre

### Reactome

```{r, fig.height=10}
gsea_plots[['RB_ind_pre-v-12_reactome']] 
```

### Hallmark & KEGG

```{r, fig.height=10}
gsea_plots[['RB_ind_pre-v-12_hallmark&kegg']]
```

## 18 v pre

### Reactome

```{r, fig.height=10}
gsea_plots[['RB_ind_pre-v-18_reactome']]
```

### Hallmark & KEGG

```{r, fig.height=6}
gsea_plots[['RB_ind_pre-v-18_hallmark&kegg']]
```

## 18 v 12

### Reactome

```{r, fig.height=10}
gsea_plots[['RB_ind_18-v-12_reactome']]
```

### Hallmark & KEGG

```{r, fig.height=6}
gsea_plots[['RB_ind_18-v-12_hallmark&kegg']]
```

## Joint

### Reactome

```{r, fig.height=10}
gsea_plots[['RB_joint_reactome']]
```

### Hallmark & KEGG

```{r, fig.height=8}
gsea_plots[['RB_joint_hallmark+kegg']]
```

## Running enrichment plots {.tabset}

### HALLMARK_INFLAMMATORY_RESPONSE

```{r}
gsea_plots$RB_HALLMARK_INFLAMMATORY_RESPONSE
```

### REACTOME_INTERFERON_GAMMA_SIGNALING

```{r}
gsea_plots$RB_REACTOME_INTERFERON_GAMMA_SIGNALING
```

### REACTOME_INTERFERON_SIGNALING

```{r}
gsea_plots$RB_REACTOME_INTERFERON_SIGNALING
```

### REACTOME_SIGNALING_BY_INTERLEUKINS

```{r}
gsea_plots$RB_REACTOME_SIGNALING_BY_INTERLEUKINS
```

### REACTOME_TOLL_LIKE_RECEPTOR_CASCADES

```{r}
gsea_plots$RB_REACTOME_TOLL_LIKE_RECEPTOR_CASCADES
```

### KEGG_B_CELL_RECEPTOR_SIGNALING_PATHWAY

```{r}
gsea_plots$RB_KEGG_B_CELL_RECEPTOR_SIGNALING_PATHWAY
```
