---
title: "model_validation"
author: "ENPRC Gencore"
date:  "`r Sys.Date()`"
output: 
  rmdformats::robobook:
    fig_width: 9
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = TRUE,
                      message = TRUE, 
                      echo=FALSE,
                      fig.width = 6,
                      fig.height = 4,
                      dev='png')
library(tidyverse)    ## General logic and processing
library(DESeq2)       ## Makes us have a job
library(here)         ## For consistent directory structures
# library(edgeR)
library(DT)
library(reshape2)
library(gencoreBulk)
library(dplyr)
library(knitr)
library(kableExtra)
```

# Read metadata

```{r}
obj <- readRDS(here('rds/mnem_analysis_post_QC.Rds'))
```

```{r}
obj$dds$Timepoint <- factor(plyr::mapvalues(obj$dds$Timepoint, 
                                            from=c(1,2,3),
                                            to=c('pre', 'W12', 'W18')), 
                            levels=c('pre', 'W12', 'W18'))
```

```{r}
metadata <- as.data.frame(obj$dds@colData)
exp_data <- obj$dds
plots <- list()
```

```{r}
metadata %>%
  select(-c(FileID, SampleID, sizeFactor, replaceable))
  # select(sampleID, SubjectID, grp)
```

```{r}
relevant_cols <- colnames(metadata %>% select(-c(FileID, SampleID, sizeFactor, replaceable)))
metadata_summary <- lapply(relevant_cols, function(x) table(unlist((metadata[[x]]))))
names(metadata_summary) <- relevant_cols
metadata_summary
```

```{r}
table(metadata$Individual, metadata$Timepoint)
```

```{r}
table(metadata$Individual, metadata$Timepoint, metadata$Tissue)
```

# Design 1 - Combined tissues

```{r}
design_formula <- as.formula('~ Tissue + Timepoint + Individual + Timepoint:Tissue')
design_matrix  <- stats::model.matrix(design_formula, data = metadata)
colnames(design_matrix)
```

```{r}
design_matrix
```

## Fit design

```{r}
design(obj$dds) <- design_formula
stopifnot(colnames(obj$dds) == rownames(obj$dds@colData))
obj$dds <- DESeq2::DESeq(obj$dds)
```


## Define contrasts

Explicitly write out full contrast formulas, including terms that will cancel and
expanded interaction terms.
This avoids mistakes and facilitates plotting.

You do not have to use valid R names at this stage, as that will be corrected
for later. Just use the colnames of the design matrix.

```{r}
colnames(design_matrix)
```

Use spaces around terms to make sure they are detected by term extraction
functions for plotting (WIP)

```{r}
contrast_matrix_expanded <- data.frame(
  contrast_names = c(
    'W12_v_pre.allTissue',
    'W12_v_pre.rb',
    'W12_v_pre.pbmc',
    'W18_v_pre.allTissue',
    'W18_v_pre.rb',
    'W18_v_pre.pbmc',
    'W18_v_W12.allTissue',
    'W18_v_W12.rb',
    'W18_v_W12.pbmc',
    'rb_v_pbmc.W12',
    'rb_v_pbmc.W18'
  ),
  numerators = c(
   '(Intercept) + TimepointW12 + 0.5*( Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW12 )',
   '(Intercept) + TimepointW12 + Tissuerectal-biopsy+ Tissuerectal-biopsy:TimepointW12',
   '(Intercept) + TimepointW12',
   '(Intercept) + TimepointW18 + 0.5*( Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW18 )',
   '(Intercept) + TimepointW18 + Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW18',
   '(Intercept) + TimepointW18',
   '(Intercept) + TimepointW18 + 0.5*( Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW18 )',
   '(Intercept) + TimepointW18 + Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW18',
   '(Intercept) + TimepointW18',
   '(Intercept) + TimepointW12 + Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW12',
   '(Intercept) + TimepointW18 + Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW18'
  ),
  denominators = c(
   '(Intercept) + 0.5* Tissuerectal-biopsy',
   '(Intercept) + Tissuerectal-biopsy',
   '(Intercept)',
   '(Intercept) + 0.5* Tissuerectal-biopsy',
   '(Intercept) + Tissuerectal-biopsy',
   '(Intercept)',
   '(Intercept) + TimepointW12 + 0.5*( Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW12 )',
   '(Intercept) + TimepointW12 + Tissuerectal-biopsy + Tissuerectal-biopsy:TimepointW12',
   '(Intercept) + TimepointW12',
   '(Intercept) + TimepointW12',
   '(Intercept) + TimepointW18'
  )
)
```


```{r}
contrast_matrix <- contrast_matrix_expanded
for (term in rev(colnames(design_matrix))) {
  contrast_matrix <- contrast_matrix %>%
    mutate(numerators = gsub(term, replacement = make.names(term), numerators, fixed = TRUE),
           denominators = gsub(term, replacement = make.names(term), denominators, fixed = TRUE))
}

## Then make contrasts
contrast_matrix <- apply(contrast_matrix, MARGIN = 1, function(row) {
  paste0('(', row[2], ') - (', row[3], ')')
})
contrast_matrix <- limma::makeContrasts(contrasts = contrast_matrix,
    levels = make.names(colnames(design_matrix)))
colnames(contrast_matrix) <- contrast_matrix_expanded$contrast_names
contrast_matrix
```

## Extract contrasts

### DESeq

```{r}
model_results <- apply(contrast_matrix, 2, gencoreBulk::extractFilteredDESeqResults, dds = obj$dds)
names(model_results) <- colnames(contrast_matrix)
```

## Check results

```{r}
lapply(model_results, summary)
```

### Summary table

```{r}
## Not very generalized yet
data.table::rbindlist(lapply(model_results, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
```

## Visualize 

Plot expression boxplots for quick inspections

```{r, fig.height=10, fig.width=7}
genes <- getTopNGenes(model_results$W18_v_pre.rb, exclude_ENS = TRUE, ENS_pattern = 'ENSMNEG')
data_to_plot = assays(obj$dds)$rld[,assays(obj$dds)$rld$Tissue == 'rectal-biopsy']

data_to_plot <- normalizeCountsForHeatmapByIndividual(assay(data_to_plot),
                                   data_to_plot@colData,
                                   group_var = 'Timepoint', baseline = '1', 
                                   individual_var = 'Individual',
                                   remove_baseline = TRUE)
  
heatmapFromGenelist(genes,
                    data_to_plot)
```

```{r}
model_results$W18_v_pre.rb %>%
  as.data.frame() %>%
  filter(row.names(.) %in% genes)
```

# Design 2 - Separate tissues

```{r}
design_formula <- as.formula('~ Timepoint + Individual')
design_matrix  <- stats::model.matrix(design_formula, data = metadata)
colnames(design_matrix)
```

```{r}
design_matrix
```

## Fit design

```{r}
obj.rb <- obj$dds[,obj$dds$Tissue=='rectal-biopsy']
obj.pbmc <- obj$dds[,obj$dds$Tissue=='PBMC']

design(obj.rb) <- design_formula
stopifnot(colnames(obj.rb) == rownames(obj.rb@colData))
obj.rb <- DESeq2::DESeq(obj.rb)

design(obj.pbmc) <- design_formula
stopifnot(colnames(obj.pbmc) == rownames(obj.pbmc@colData))
obj.pbmc <- DESeq2::DESeq(obj.pbmc)
```


## Define contrasts

Explicitly write out full contrast formulas, including terms that will cancel and
expanded interaction terms.
This avoids mistakes and facilitates plotting.

You do not have to use valid R names at this stage, as that will be corrected
for later. Just use the colnames of the design matrix.

```{r}
colnames(design_matrix)
```

Use spaces around terms to make sure they are detected by term extraction
functions for plotting (WIP)

```{r}
contrast_matrix_expanded <- data.frame(
  contrast_names = c(
    'W12_v_pre',
    'W18_v_pre',
    'W18_v_W12'
  ),
  numerators = c(
   '(Intercept) + TimepointW12',
   '(Intercept) + TimepointW18',
   '(Intercept) + TimepointW18'
  ),
  denominators = c(
   '(Intercept)',
   '(Intercept)',
   '(Intercept) + TimepointW12'
  )
)
```


```{r}
contrast_matrix <- contrast_matrix_expanded
for (term in rev(colnames(design_matrix))) {
  contrast_matrix <- contrast_matrix %>%
    mutate(numerators = gsub(term, replacement = make.names(term), numerators, fixed = TRUE),
           denominators = gsub(term, replacement = make.names(term), denominators, fixed = TRUE))
}

## Then make contrasts
contrast_matrix <- apply(contrast_matrix, MARGIN = 1, function(row) {
  paste0('(', row[2], ') - (', row[3], ')')
})
contrast_matrix <- limma::makeContrasts(contrasts = contrast_matrix,
    levels = make.names(colnames(design_matrix)))
colnames(contrast_matrix) <- contrast_matrix_expanded$contrast_names
contrast_matrix
```

## Extract contrasts

### DESeq

```{r}
model_results.rb <- apply(contrast_matrix, 2, gencoreBulk::extractFilteredDESeqResults, dds = obj.rb)
names(model_results.rb) <- colnames(contrast_matrix)

model_results.pbmc <- apply(contrast_matrix, 2, gencoreBulk::extractFilteredDESeqResults, dds = obj.pbmc)
names(model_results.pbmc) <- colnames(contrast_matrix)
```

## Check results

```{r}
lapply(model_results.rb, summary)
lapply(model_results.pbmc, summary)
```

### Summary table

```{r}
## Not very generalized yet
data.table::rbindlist(lapply(model_results, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
```

## Visualize 

Plot expression boxplots for quick inspections

```{r, fig.height=10, fig.width=7}
genes <- getTopNGenes(model_results.rb$W12_v_pre, 
                      exclude_ENS = TRUE, ENS_pattern = 'ENSMNEG')
data_to_plot <- normalizeCountsForHeatmapByIndividual(assay(assays(obj.rb)$rld),
                                   obj.rb@colData,
                                   group_var = 'Timepoint', baseline = 'pre', 
                                   individual_var = 'Individual',
                                   remove_baseline = TRUE)
  
heatmapFromGenelist(genes,
                    data_to_plot)
```

# Compare approaches

```{r}
model_results.rb$W12_v_pre %>%
  as.data.frame() %>%
  rownames_to_column('gene') %>%
  filter(gene %in% genes)
```

```{r}
tmp <- data.table::rbindlist(lapply(model_results.rb, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
tmp$contrast <- paste0(names(model_results.rb), '.rb')

tmp1 <- data.table::rbindlist(lapply(model_results.pbmc, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
tmp1$contrast <- paste0(names(model_results.pbmc), '.pbmc')

tmp1 <- rbind(tmp, tmp1)

tmp <- data.table::rbindlist(lapply(model_results, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
tmp$contrast <- names(model_results)

model_comp_table <- merge(tmp1, tmp, by='contrast', suffixes=c('.separate', '.joined'), all=TRUE) %>%
  select(up.separate, down.separate, contrast, up.joined, down.joined) %>%
  mutate(contrast=factor(contrast, levels = c(
    "W12_v_pre.pbmc", "W12_v_pre.rb", "W12_v_pre.allTissue", 
    "W18_v_pre.pbmc", "W18_v_pre.rb", "W18_v_pre.allTissue",
    "W18_v_W12.pbmc", "W18_v_W12.rb", "W18_v_W12.allTissue", 
    "rb_v_pbmc.W12", "rb_v_pbmc.W18" 
  ))) %>%
  arrange(contrast)

model_comp_table <- kable(model_comp_table, col.names = c('Up', 'Down', 'Contrast', 'Up', 'Down'),
             align = 'c') %>%
  kable_classic() %>%
  add_header_above(c("Separate" = 2, " " = 1,  "Joined" = 2)) %>%
  kable_styling("striped")
model_comp_table
```


```{r}
tmp <- data.table::rbindlist(lapply(model_results.rb, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
tmp$contrast <- names(model_results.rb)

tmp1 <- data.table::rbindlist(lapply(model_results.pbmc, function(x) {
  up <- x %>% as.data.frame() %>% filter(log2FoldChange > 0) %>% filter(padj<0.05) %>% nrow()
  down <- x %>% as.data.frame() %>% filter(log2FoldChange < 0) %>% filter(padj<0.05) %>% nrow()
  return(list(up=up, down=down))
}))
tmp1$contrast <- names(model_results.pbmc)

model_summary_table <- merge(tmp, tmp1, by = 'contrast', suffixes = c('.rb', '.pbmc')) %>%
  select(up.rb, down.rb, contrast, up.pbmc, down.pbmc) %>%
  mutate(contrast=factor(contrast, levels = c(
    "W12_v_pre",
    "W18_v_pre",
    "W18_v_W12"
  ))) %>%
  arrange(contrast)

model_summary_table <- kable(model_summary_table, 
                             col.names = c('Up', 'Down', 'Contrast', 'Up', 'Down'),
                             align = 'c') %>%
  kable_classic() %>%
  add_header_above(c("Rectal biopsy" = 2, " " = 1,  "PBMC" = 2)) %>%
  kable_styling("striped")
model_summary_table
```

# Case study

```{r}
tmp <- unlist(lapply(model_results.pbmc, function(x) {
  as.data.frame(x) %>% 
    rownames_to_column('gene') %>%
    filter(padj < 0.05) %>%
    select(gene)
}))
tmp <- table(tmp)
tmp[tmp==1]
```


```{r}
genes <- c('CD19', 'CD22', 'FCRL2', 'CHIT1', 'TCN2')
data_to_plot <- normalizeCountsForHeatmapByIndividual(assay(assays(obj.pbmc)$rld),
                                   obj.pbmc@colData,
                                   group_var = 'Timepoint', baseline = 'pre', 
                                   individual_var = 'Individual',
                                   remove_baseline = TRUE)
casestudy_heatmap_pbmc <- heatmapFromGenelist(genes, data_to_plot,
                    column_split = c(rep(1,5), rep(2,5)), 
                    slice_labels = c('W12', 'W18'), 
                    slice_labels_rot = 0, 
                    column_labels = gsub('.+(Z[0-9]+).+', '\\1',
                                         colnames(data_to_plot)))
casestudy_heatmap_pbmc
```

```{r}
casestudy_table <- merge(
  as.data.frame(model_results.pbmc$W12_v_pre) %>%
    rownames_to_column('gene') %>%
    filter(gene %in% genes) %>%
    select(gene, log2FoldChange, lfcSE, padj),
  as.data.frame(model_results.pbmc$W18_v_pre) %>%
    rownames_to_column('gene') %>%
    filter(gene %in% genes) %>%
  select(gene, log2FoldChange, lfcSE, padj),
  by = 'gene', suffixes = c('.W12', '.W18')
)

casestudy_table <- merge(
  casestudy_table, 
  as.data.frame(model_results.pbmc$W18_v_W12) %>%
    rownames_to_column('gene') %>%
    filter(gene %in% genes) %>%
  select(gene, log2FoldChange, lfcSE, padj),
  by = 'gene', suffixes = c('', '.W18vW12')
) %>%
  mutate(gene = factor(gene, levels = genes)) %>%
  arrange(gene)

casestudy_table <- data.frame(lapply(casestudy_table, function(y) if(is.numeric(y)) round(y, 3) else y)) 

casestudy_table <- kable(casestudy_table, 
                             col.names = c('Gene',rep(c('LFC', 'LFC SE', 'adj. P') ,3)),
                             align = 'c',
                         caption = 'LFC = log2 fold change; SE = standard error; adj. P = adjusted P value') %>%
  kable_classic() %>%
  add_header_above(c(" " = 1, "W12" = 3, "W18" = 3, "W18_v_W12" =3 )) %>%
  kable_styling("striped") %>%
  column_spec (c(2,5,8),border_left = T, border_right = F) 
casestudy_table
```

# Render

```{r}
rmarkdown::render(here::here('scripts/model_validation.format.Rmd'),
                  output_file = 'model_validation.html',
                  output_dir = here::here('reports'),
                  params = list(
                    title = 'Bulk RNA Analysis Report')
                  )
```

# Save model

```{r}
saveRDS(obj.pbmc, here('rds/PBMC/pbmc_obj-post_model_fit.Rds'))
saveRDS(obj.rb, here('rds/RB/rb_obj-post_model_fit.Rds'))

saveRDS(model_results.pbmc, here('rds/PBMC/PBMC-extracted_model_fits.Rds'))
saveRDS(model_results.rb, here('rds/RB/RB-extracted_model_fits.Rds'))
```

